---
title: "iButton Temperature Data from 2021-2023 NIRPO Plots in Prudhoe Bay, Alaska"
author: "Jana Peirce"
date: "6 November 2023"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)
library(here)
library(stringr)
library(tinytex)
library(readxl)
library(lubridate)
library(hms)
```

## File prep

- Save all data with datetime info in csv before bringing into r (do not use readxl package)
- Format all times as yyyy-mm-dd before importing into r

```{r}

# Read the CSV file
ibtn_temp2 <- read.csv("data/ibtn_data_1921G.csv")

```

```{r}

# view output

head(ibtn_temp2)

```

```{r}
# transform data table from wide to long format

temp_long2 <- ibtn_temp2 %>% 
  pivot_longer(cols = c(-Date, -Time), names_to = "iBtn_ID", values_to = "temp_c") 

# view output

head(temp_long2)

```

```{r}
# convert date col from char to date format (To work must start with date in YYYY-MM-DD in Excel file first)
date_temp_long2X <- data.frame(temp_long2)
  mutate(temp_long2, Date= as.Date(Date, format= "%Y-%m-%d"))

```


```{r}
# remove leading X from iBtn_ID

date_temp_long2 <- date_temp_long2X %>% 
  mutate_at("iBtn_ID", str_replace, "X", "")

knitr::kable(head(date_temp_long2[, 1:4]), "simple", col.names = c("Date","Time","iBtn ID","Temp (\u00B0C)"))
```

```{r}
#read in plot type table

plot_type2 <- read.csv("data/ibtn_summary.csv")
```

```{r}

# view output
head(plot_type2)

```

```{r}

# join both tables together
ibtn_summary2X <- left_join(date_temp_long2, plot_type2, by = c("iBtn_ID")) 

# view output
head(ibtn_summary2X)

```

```{r}
# NOT SURE IF THIS IS NEEDED - convert date col AGAIN from char to date format
ibtn_summary2 <- data.frame(ibtn_summary2X)
  mutate(ibtn_summary2X, Date= as.Date(Date, format= "%Y-%m-%d"))

# view output - Date column is still back to chr in short df, but shows as date in full data.frame
head(ibtn_summary2)
```

```{r}
#read in plot type table

plot_var2 <- read.csv("data/plot_enviro_data.csv")

```

```{r}

# view output
head(plot_var2)

```

```{r}

# join both tables together
ibtn_summary2all <- left_join(ibtn_summary2, plot_var2, by = c("Plot_ID")) 

# view output - why is date back to being a char field? Is it really? See tibble vs data frame?
head(ibtn_summary2all)

```
```{r}
write.csv(ibtn_summary2all, "data/ibtn_summary2all.csv")

```

### Average Daily Temperature calculated (4320 rows)

```{r}
# calculate average daily temp for each date by iButton

avg_daily_temp2 <- ibtn_summary2all %>%
  group_by(Date, iBtn_ID, Plot_ID, Veg_type, Depth, Transect.x, Moist_grad) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop") %>% 
  mutate(across(where(is.numeric), ~ round(., 1)))

knitr::kable(head(avg_daily_temp2[, 1:8]), "simple", col.names = c("Date","iBtn ID","Plot ID","Veg Type","Depth","Transect","Moisture","Avg Daily Temp (\u00B0C)"))
```

```{r}
write.csv(avg_daily_temp2, "data/adt_ibtn2.csv")

```


```{r}
# calculate avg daily temp for each date by depth for each veg type

adt_depthXvegtype2 <- ibtn_summary2all %>% 
  group_by(Date, Depth, Veg_type) %>% 
  summarize(avg_daily_temp2 = mean(temp_c, na.rm = T), .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

knitr::kable(head(adt_depthXvegtype2[, 1:4]), "simple", col.names = c("Date","Depth","Veg Type","Avg Daily Temp (\u00B0C)"))
```


```{r}
write.csv(adt_depthXvegtype2, "data/adt_depthXvegtype2.csv")

```


```{r}
# calculate mean for all days and times by iBtn

mean_iBtn2 <- ibtn_summary2all %>% 
  group_by(iBtn_ID, Plot_ID, Veg_type, Depth, Transect.x, Moist_grad) %>% 
  summarize(avg_daily_temp2 = mean(temp_c, na.rm = T), .groups = "drop") %>% 
  mutate(across(where(is.numeric), ~ round(., 4)))

knitr::kable(head(mean_iBtn2[, 1:7]), "simple", col.names = c("iBtn ID","Plot ID","Veg Type","Depth","Transect","Moisture","Avg Daily Temp (\u00B0C)"))

```

```{r}
write.csv(mean_iBtn2, "data/mean_iBtn2.csv")

```

```{r}
# calculate mean for all days and times for each plot at each depth there is an iBtn

mean_plotID2 <- ibtn_summary2all %>% 
  group_by(Plot_ID, Veg_type, Depth, Transect.x, Moist_grad) %>%
  summarize(avg_daily_temp2 = mean(temp_c, na.rm = T), .groups = "drop") %>% 
  mutate(across(where(is.numeric), ~ round(., 4)))

knitr::kable(head(mean_plotID2[, 1:6]), "simple", col.names = c("Plot ID","Depth","Veg Type","Transect","Moisture","Avg Daily Temp (\u00B0C)"))

```

```{r}
write.csv(mean_plotID2, "data/mean_plotID2.csv")

### STOPPED HERE FOR NOW - need to calculate for each plot at all depths and veg types - not grouped. Need to make a df for just surface (depth = 0) ibuttons.
```

```{r}
# plot mean veg type - need to restrict to depth = 0

p <- ggplot(mean_plotID2, aes(Veg_type, avg_daily_temp2)) +
  geom_col(fill="salmon") +
  labs(title = "Mean Temperature - All Days - by Veg type", x = "Veg Type", y = "Mean Temp (\u00B0C)") + theme(panel.background = element_blank(), plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = avg_daily_temp2), vjust = -0.5)

p

knitr::kable(head(mean_plotID2[, 1:2]), "simple", col.names = c("Veg Type", "Mean Temp (\u00B0C)"))

p.mean_plot <- p
```
```{r}
# calculate avg daily temp by ibtn position

adt_location <- ibtn_summary %>% 
  group_by(Date, Location) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop")
```

```{r}
# calculate mean for all days by ibtn position

mean_location <- ibtn_summary %>% 
  group_by(Location) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop") %>% 
  mutate(across(where(is.numeric), ~ round(., 1)))
```

```{r}
# plot mean for all ibtns by position

p <- ggplot(mean_location, aes(Location, avg_daily_temp)) +
  geom_col(fill="salmon") +
  labs(title = "Mean Temperature - All Days - by iBtn Position", x = "iBtn Position", y = "Mean Temp (\u00B0C)") + theme(panel.background = element_blank(), plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = avg_daily_temp), vjust = -0.5)

p

knitr::kable(head(mean_veg_type[, 1:2]), "simple", col.names = c("iBtn Position", "Mean Temp (\u00B0C)"))

p.mean_sensor <- p
```

```{r}
# subset to remove sensors not assoc with plots
# then calculate average daily temp for each day by plot type

ibtn_summary_geol <- subset(ibtn_summary, Surf_geol!="N/A")

adt_surf_geol <- ibtn_summary_geol %>% 
  group_by(Date, Surf_geol) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop")
```

```{r}
# calculate mean for all days by plot type

mean_surf_geol <- ibtn_summary_geol %>% 
  group_by(Surf_geol) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop") %>% mutate(across(where(is.numeric), ~ round(., 1)))
```

```{r}
# plot mean plot type for season

p <- ggplot(mean_surf_geol, aes(Surf_geol, avg_daily_temp)) +
  geom_col(fill="darkolivegreen3") +
  labs(title = "Mean Temperature by Surficial Geology", x = "Surf geol", y = "Mean Temp (\u00B0C)") + theme(panel.background = element_blank(), plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = avg_daily_temp), vjust = -0.5)

p

knitr::kable(head(mean_surf_geol[, 1:2]), "simple", col.names = c("Surficial Geology", "Mean Temp (\u00B0C)"))

p.mean_plot <- p
```


```{r}
# calculate mean for all days by veg type and surficial geology

mean_veg_geol <- ibtn_summary_geol %>% 
  group_by(Veg_type, Surf_geol) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop") %>% 
  mutate(across(where(is.numeric), ~ round(., 1)))

p <- ggplot(mean_veg_geol, aes(x = Veg_type, y = avg_daily_temp, fill = Surf_geol)) +
  geom_col(position = "dodge2") +
  labs(title = "Mean Temperature by Veg Type and Surface Geology", x = NULL, y = "Mean Temp (\u00B0C)") + theme(panel.background = element_blank(), plot.title = element_text(hjust = 0.5), legend.title = element_blank()) 

p

knitr::kable(mean_veg_geol[, 1:3], "simple", col.names = c("Veg Type","Surf Geol","Mean Temp (\u00B0C)"))

p.mean_veg_geol <- p
```
```{r}
# calculate mean for all days by veg type and iBtn position

mean_veg_loc <- ibtn_summary %>% 
  group_by(Veg_type, Location) %>% 
  summarize(avg_daily_temp = mean(temp_c, na.rm = T), .groups = "drop") %>% 
  mutate(across(where(is.numeric), ~ round(., 1)))

p <- ggplot(mean_veg_loc, aes(x = Veg_type, y = avg_daily_temp, fill = Location)) +
  geom_col(position = "dodge2") +
  labs(title = "Mean Temperature by Veg Type and iBtn Position", x = NULL, y = "Mean Temp (\u00B0C)") + theme(panel.background = element_blank(), plot.title = element_text(hjust = 0.5), legend.title = element_blank()) 

p

knitr::kable(mean_veg_geol[, 1:3], "simple", col.names = c("Veg Type","iBtn Position","Mean Temp (\u00B0C)"))

p.mean_veg_loc <- p
```

```{r}
#plot adt all ibuttons

p <- ggplot(avg_daily_temp, aes(x = Date, y = avg_daily_temp)) +
  geom_point(alpha = 0.3, color = "dodgerblue2") +
  labs(title = "Average Daily Temperature", x = NULL, y = "Temperature (\u00B0C)") +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_all_ibtn <- p
```

### Let's look at the data in more detail. 

Here is a plot of the average daily temperature of each iButton from all vegetation types at soil surface and the base of the organic layer.

```{r}
# plot by iBtn location

p <- ggplot(avg_daily_temp, aes(x = Date, y = avg_daily_temp, color=Location)) +
  geom_point(alpha = 0.3) +
  labs(title = "Average Daily Temperature by iBtn Position", x = NULL, y = "Temperature (\u00B0C)", color = "iButton Position")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_ibtn_loc <-  p
```
Now we can look at the just the surface data colored to indicated the vegetation type of the plot.

```{r}
# subset to remove sensors at the base of the organic layer

adt_veg_surface <- subset(avg_daily_temp, Location!="base org layer")
```

```{r}
# plot by Veg type at soil surface

p <- ggplot(adt_veg_surface, aes(x = Date, y = avg_daily_temp, color=Veg_type)) +
  geom_point(alpha = 0.5) +
  labs(title = "Average Daily Temperature at Soil Surface by Veg Type", x = NULL, y = "Temperature (\u00B0C)", color = "Location")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_veg_surface <- p

```
```{r}
# plot by Transect

p <- ggplot(adt_veg_surface, aes(x = Date, y = avg_daily_temp, color=Transect.x)) +
  geom_point(alpha = 0.3) +
  labs(title = "Average Daily Temperature by Transect at Soil Surface", x = NULL, y = "Temperature (\u00B0C)", color = "Transect")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_trans_surface <-  p
```
Now we can look at the just the base of the organic layer data colored to indicated the vegetation type of the plot.

```{r}
# subset to remove sensors at the soil surface

adt_veg_base <- subset(avg_daily_temp, Location!="surface")
```

```{r}
# plot by Veg type at base of org layer

p <- ggplot(adt_veg_base, aes(x = Date, y = avg_daily_temp, color=Veg_type)) +
  geom_point(alpha = 0.5) +
  labs(title = "Average Daily Temperature at Base of Organic Layer by Veg Type", x = NULL, y = "Temperature (\u00B0C)", color = "Location")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_veg_base <- p

```

Here is a plot of the average daily temperature of each iButton by site moisture

```{r}
# plot by moisture gradient

p <- ggplot(adt_veg_surface, aes(x = Date, y = avg_daily_temp, color=Moist_grad)) +
  geom_point(alpha = 0.3) +
  labs(title = "Average Daily Temperature at Soil Surface by Site Moisture", x = NULL, y = "Temperature (\u00B0C)", color = "Site Moisture")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_surf_moisture <-  p
```

```{r}
# plot by moisture gradient

p <- ggplot(adt_veg_base, aes(x = Date, y = avg_daily_temp, color=Moist_grad)) +
  geom_point(alpha = 0.3) +
  labs(title = "Average Daily Temperature at Base of the Organic Layer by Site Moisture", x = NULL, y = "Temperature (\u00B0C)", color = "Site Moisture")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_base_moisture <-  p
```

```{r}
# Facet plot by veg type

p <- ggplot(adt_veg_surface, aes(x = Date, y = avg_daily_temp)) +
  geom_point(alpha=0.2, color = "firebrick") +
  facet_wrap(~Veg_type, ncol=2) +
  labs(title = "Average Daily Temperature by Veg type", x = NULL, y = "Temp (\u00B0C)")  +
  theme(plot.title = element_text(hjust = 0.5))

p

p.adt_veg_facet <- p
```
```{r}
# filter out some plots for smaller facet set

transect_filter <- avg_daily_temp %>% 
    filter(Transect.x == "T6", Location == "surface")
```

```{r}
# Facet plots by select plots

p <- ggplot(transect_filter, aes(x = Date, y = avg_daily_temp)) +
  geom_point(alpha = 0.3, color = "hotpink2") +
  facet_wrap(~Plot_ID, ncol = 4) + 
  labs(title = "Avg. Daily Temp at Soil Surface in T6 Plots", x = NULL, y = "Temperature (\u00B0C)") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle=45, hjust = 1))

p

p.t6_surface_facet <- p

```

### That's it! (for now)
